Retrofitting Effect Handlers onto OCaml

### 2020-11-23 retro-concurrency.tex
Minor edits

We can modularly extend the calculator to read user input using the |Read|
effect defined earlier as follows:
\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string (perform (Read stdin))
\end{lstlisting}
The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:
\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
		register_async_input_line (chan, k);
		continue (get_ui_k ()) ()
\end{lstlisting}
\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither of the handlers needs to be aware of the other.

\subsection{Retrofitting effect handlers}

@@

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to increased
memory allocations as continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless and
hence, we would lose compatibility with tools that inspect the program stack.
Hence, we choose not to use CPS translation and represent continuations as call
stacks.

@@

\noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program does
not break under an OCaml that is extended with effect handlers. OCaml code that does not
use effect handlers will pay negligible performance and memory cost.

@@

The remainder of the paper continues with a description of the stock OCaml
program stack (\S\ref{sec:stack}). We then refine the design further by
focussing on retrofitting effect handlers into a mainstream \emph{systems}
language (\S\ref{sec:refine}), followed by the compilation scheme and
operational semantics for Multicore OCaml effect handlers
(\S\ref{sec:semantics}).  We then discuss the compiler and runtime system
support for implementing effect handlers (\S\ref{sec:impl}), and present an
extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
our design goals (\S\ref{sec:req}). Finally, we discuss the related work
(\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with End_of_file -> close_in ic; close_out oc
	   | e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end
of input, which is handled by the exception handler which closes the channels.
The |close_*| functions do nothing if the channel is already closed. The code
is written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the catch-all
exception handler closes the channels and reraises the exception to communicate
the exceptional behaviour to the caller.

@@

We can then use a top-level I/O handler similar to the introductory example
(\S\ref{sec:aio}) to discharge the I/O operations asynchronously and resumes
with the result. This handles value return cases, but what about the
exceptional cases |End_of_file| and |Sys_error|? We introduce a |discontinue|
primitive to resume a continuation by raising an exception. In this example, on
reaching the end of file, we would discontinue the continuation of the |copy|
function with |discontinue k End_of_file|, which raises the exception at
|input_line|, and the open channels will be closed.

@@

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety, but only guarantee type safety. Programs without matching effect
handlers are well-typed Multicore OCaml programs. As a result, our static
semantics is simpler than languages that ensure effect
safety~\cite{Eff,Hillerstrom20,Leijen14,Effekt,Frank,Biernacki19}. This is
important for backwards compatibility as our goal is to retrofit effect
handlers to a language with large legacy codebases. Programs that do not use
effects remain well-typed, and those that do compose well with those that
don't.

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b) continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns an |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect an |'a| value for
resumption and return a |'b| value. The continuations may be continued with a
suitably typed value or discontinued with an exception.

@@

We present an operational semantics for a core language of effect handlers that
faithfully captures the semantics of the Multicore OCaml implementation. An
executable version of the semantics implemented as an OCaml interpreter along
with examples is included in the supplementary material.

@@

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract
machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
difference from Hillerstrom et al. is that our stacks are composed of
alternating sequence of OCaml and C stack segments. The program state is
captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
the current term $\tau$ under evaluation, its environment $\env$ and the
current stack $\stack$. The term is either an expression $e$ or a value $v$.
The values are integer constants $n$, continuations $k$, closures
$\clos{x}{e}{\env}$, effects being performed ($\effval{l}{k}$) and exceptions
being raised ($\exnval{l}$). The environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the current
continuation $k$ and the C stack $\cstack$ under it. Thus, the program stack is
an alternating sequence of C and OCaml stacks terminating with an empty OCaml
stack $\ostackemp$. The frame list $\fl$ is composed of individual frames $r$,
which is one of an argument frame $\farg{e}{\env}$ with the expression $e$ at
the argument position of an application with its environment $\env$, a function
frame $\ffun{v}$ with the value $v$ at the function position of an application,
and frames for evaluating the arguments of an arithmetic expression.

Continuations $k$ are either empty $[]$ or a non-empty list of \emph{fibers}. A
fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
environment $\env$. We use the infix operator $@$ for appending two lists.

>>>>

We can modularly extend the calculator to read user input using the |Read|
effect defined earlier as follows:
\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string (perform (Read stdin))
\end{lstlisting}
The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:
\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
		register_async_input_line (chan, k);
		continue (get_ui_k ()) ()
\end{lstlisting}
\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither the division-by-zero handler nor the I/O handler needs to be aware
of the other.

\vspace{-1.5mm}
\subsection{Retrofitting effect handlers}

@@

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to the increased
memory allocations as the continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless and
hence, we would lose compatibility with tools that inspect the program stack.
Hence, we choose not to use CPS translation and represent the continuations as
call stacks.

@@

\noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program
does not break under OCaml extended with effect handlers. OCaml code that does
not use effect handlers will pay negligible performance and memory cost.

@@

The remainder of the paper continues with a description of the stock OCaml
program stack (\S\ref{sec:stack}). We then refine the design further by
focussing on retrofitting effect handlers into a mainstream \emph{systems}
language (\S\ref{sec:refine}), followed by the compilation scheme and
operational semantics for Multicore OCaml effect handlers
(\S\ref{sec:semantics}).  We then discuss the compiler and the runtime system
support for implementing effect handlers (\S\ref{sec:impl}), and present an
extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
our design goals (\S\ref{sec:req}). Finally, we discuss the related work
(\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17, Leijen18}. Consider the following function that uses
blocking I/O functions from the OCaml standard library to copy data from the
input channel |ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with End_of_file -> close_in ic; close_out oc
	   | e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end
of input, which is handled by the exception handler which closes the channels.
The |close_*| functions do nothing if the channel is already closed. The code
is written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the catch-all
exception handler closes the channels and reraises the exception to communicate
the exceptional behaviour to the caller.

@@

We can then use a top-level I/O handler, similar to the introductory example
(\S\ref{sec:aio}), to discharge the I/O operations asynchronously and resume
with the result. This handles value return cases, but what about the
exceptional cases |End_of_file| and |Sys_error|? To this end, we introduce a
|discontinue| primitive to resume a continuation by raising an exception. In
this example, on reaching the end of file, we would discontinue the captured
continuation of the |input_line| function with |discontinue k End_of_file|,
which raises the exception at |input_line|, and the open channels will be
closed.

@@

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety, but only guarantee type safety. Programs without matching effect
handlers are well-typed Multicore OCaml programs. As a result, our static
semantics is simpler than languages that ensure effect safety~\cite{Eff,
Hillerstrom20, Leijen14, Effekt, Frank, Biernacki19}. This is important for
backwards compatibility as our goal is to retrofit effect handlers to a
language with large legacy codebases; programs that do not use effects remain
well-typed, and those that do compose well with those that don't.

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b) continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns an |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expects a |'a| value for
resumption and returns a |'b| value. The continuations may be continued with a
suitably typed value or discontinued with an exception.

@@

We present an operational semantics for a core language of effect handlers that
faithfully captures the semantics of the Multicore OCaml implementation. An
executable version of the semantics, implemented as an OCaml interpreter, along
with examples, is included in the supplementary material.

@@

The operational semantics is an extension of the CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract
machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
difference from Hillerstrom et al. is that our stacks are composed of
alternating sequence of OCaml and C stack segments. The program state is
captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
the current term $\tau$ under evaluation, its environment $\env$ and the
current stack $\stack$. The term is either an expression $e$ or a value $v$.
The values are integer constants $n$, continuations $k$, closures
$\clos{x}{e}{\env}$, effects being performed ($\effval{l}{k}$) and exceptions
being raised ($\exnval{l}$). The environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). The C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and the OCaml stack $\ostack$ under it. The OCaml stack is
either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the current
continuation $k$ and the C stack $\cstack$ under it. Thus, the program stack is
an alternating sequence of C and OCaml stacks terminating with an empty OCaml
stack $\ostackemp$. The frame list $\fl$ is composed of individual frames $r$,
which is one of an argument frame $\farg{e}{\env}$ with the expression $e$ at
the argument position of an application with its environment $\env$, a function
frame $\ffun{v}$ with the value $v$ at the function position of an application,
and frames for evaluating the arguments of an arithmetic expression.

A continuation $k$ is either empty $[]$ or a non-empty list of \emph{fibers}. A
fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
environment $\env$. We use the infix operator $@$ for appending two lists.
